---
title: "Spam Email Classifier"
author: "Justin Kim"
date: "2025-06-07"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Setup & data loading

```{r load-data}
library(ggplot2)
library(gridExtra)
library(MASS)
library(e1071)

# load in train and test sets
train_raw <- read.table("data/spam-train.txt", sep = ",", header = FALSE)
test_raw  <- read.table("data/spam-test.txt",  sep = ",", header = FALSE)

# cols 1-57 = features, col 58 = spam label (0 or 1)
X_train <- train_raw[, 1:57]
y_train <- train_raw[, 58]
X_test  <- test_raw[, 1:57]
y_test  <- test_raw[, 58]

# convert labels to factors for classification functions
y_train_fac <- factor(y_train, levels = c(0, 1), labels = c("ham", "spam"))
y_test_fac  <- factor(y_test,  levels = c(0, 1), labels = c("ham", "spam"))
```

## Preprocessing

I'm trying three different transformations on the raw features to see
which one gives classifiers the best shot at separating spam from ham.

### Standardize (zero mean, unit var)

```{r preprocess-std}
# important: fit stats on train only, then apply to both
col_means <- apply(X_train, 2, mean)
col_sds   <- apply(X_train, 2, sd)

X_train_std <- scale(X_train, center = col_means, scale = col_sds)
X_test_std  <- scale(X_test,  center = col_means, scale = col_sds)
```

### Log transform

```{r preprocess-log}
# log(x+1) -- the +1 keeps zeros from blowing up
X_train_log <- log(X_train + 1)
X_test_log  <- log(X_test + 1)
```

### Binarize

```{r preprocess-bin}
# just checking if the feature is nonzero
X_train_bin <- as.data.frame((X_train > 0) * 1)
X_test_bin  <- as.data.frame((X_test > 0) * 1)
```

---

## (a) PCA visualization

Reducing to 2 dimensions with PCA so I can actually see what's going on.

```{r pca-plots, fig.width=14, fig.height=5}
# standardized -- already centered/scaled so no need to redo
pca1 <- prcomp(X_train_std, center = FALSE, scale. = FALSE)
df1 <- data.frame(PC1 = pca1$x[,1], PC2 = pca1$x[,2], Label = y_train_fac)

p1 <- ggplot(df1, aes(PC1, PC2, color = Label)) +
  geom_point(alpha = 0.6, size = 1) +
  labs(title = "Standardized Features") +
  theme_minimal()

# log -- let prcomp handle centering/scaling
pca2 <- prcomp(X_train_log, center = TRUE, scale. = TRUE)
df2 <- data.frame(PC1 = pca2$x[,1], PC2 = pca2$x[,2], Label = y_train_fac)

p2 <- ggplot(df2, aes(PC1, PC2, color = Label)) +
  geom_point(alpha = 0.6, size = 1) +
  labs(title = "log(x+1) Features") +
  theme_minimal()

# binary -- center but don't scale (scaling 0/1 vars is weird)
pca3 <- prcomp(as.data.frame(lapply(X_train_bin, as.numeric)),
               center = TRUE, scale. = FALSE)
df3 <- data.frame(PC1 = pca3$x[,1], PC2 = pca3$x[,2], Label = y_train_fac)

p3 <- ggplot(df3, aes(PC1, PC2, color = Label)) +
  geom_point(alpha = 0.6, size = 1) +
  labs(title = "Binarized Features") +
  theme_minimal()

grid.arrange(p1, p2, p3, ncol = 3)
```

All three show some separation between spam and ham along the first two PCs,
but the log-transformed version seems to give the cleanest clusters. The
binarized features compress everything into a tight grid since every value
is 0 or 1.

---

## (b) Logistic regression

```{r logistic-helper}
# threshold for bonferroni correction -- testing 57 predictors
bonf_cutoff <- 0.05 / ncol(X_train_std)

# helper to fit logistic regression and return errors + significant vars
logreg <- function(xtr, ytr, xte, yte) {
  df_tr <- data.frame(xtr, y = ytr)
  df_te <- data.frame(xte, y = yte)

  fit <- glm(y ~ ., data = df_tr, family = binomial)

  # predict on both sets
  phat_tr <- predict(fit, type = "response") > 0.5
  phat_te <- predict(fit, newdata = df_te, type = "response") > 0.5

  err_tr <- mean(phat_tr != (ytr == "spam"))
  err_te <- mean(phat_te != (yte == "spam"))

  # pull out p-values
  ctab <- summary(fit)$coefficients
  pvals <- ctab[, 4]

  list(
    train_err = err_tr,
    test_err  = err_te,
    sig_005   = names(pvals)[pvals < 0.05],
    sig_bonf  = names(pvals)[pvals < bonf_cutoff],
    coef_tab  = ctab
  )
}
```

```{r logistic-fit}
lr_std <- logreg(X_train_std, y_train_fac, X_test_std, y_test_fac)
lr_log <- logreg(X_train_log, y_train_fac, X_test_log, y_test_fac)
lr_bin <- logreg(X_train_bin, y_train_fac, X_test_bin, y_test_fac)
```

### Classification errors

```{r logistic-table}
lr_errs <- data.frame(
  Preprocessing = c("Standardized", "Log-transformed", "Binarized"),
  Train = c(lr_std$train_err, lr_log$train_err, lr_bin$train_err),
  Test  = c(lr_std$test_err,  lr_log$test_err,  lr_bin$test_err)
)
knitr::kable(lr_errs, digits = 3, col.names = c("Preprocessing", "Train Err", "Test Err"))
```

### Significant predictors

```{r logistic-sig}
cat("Standardized -- p < 0.05:\n",
    paste(lr_std$sig_005, collapse = ", "), "\n\n")

cat("Standardized -- Bonferroni (p <", round(bonf_cutoff, 5), "):\n",
    paste(lr_std$sig_bonf, collapse = ", "), "\n\n")

cat("Log -- p < 0.05:\n",
    paste(lr_log$sig_005, collapse = ", "), "\n\n")

cat("Log -- Bonferroni:\n",
    paste(lr_log$sig_bonf, collapse = ", "), "\n\n")

cat("Binarized -- p < 0.05:\n",
    paste(lr_bin$sig_005, collapse = ", "), "\n\n")

cat("Binarized -- Bonferroni:\n",
    paste(lr_bin$sig_bonf, collapse = ", "), "\n")
```

Across all three, the log version had the lowest test error (~5.7%), then
standardized (~7.3%), and binarized last (~8.1%). With 57 predictors, a
bunch show up as significant at the unadjusted p < 0.05 level, but after
Bonferroni correction (α* ≈ 0.00088), only about a dozen survive. The ones
that keep showing up regardless of preprocessing are V5, V7, V16, V17, V21,
V23, V25, V27, V45, V46, V52, V53 -- these are the heavy hitters for spam
detection (word frequencies like "free", "remove", "your" plus capital letter
statistics). The binarized model also flags a few more (V11, V13, V24, V28,
V37, V42), which makes sense since for a lot of these tokens, just knowing
whether they appear at all carries most of the information.

---

## (c) LDA & QDA

Only running these on standardized and log-transformed data. Binary features
make QDA's covariance estimation unstable so I left that out.

```{r da-helper}
disc_analysis <- function(type, xtr, ytr, xte, yte) {
  df_tr <- data.frame(xtr, y = ytr)
  df_te <- data.frame(xte, y = yte)

  if (type == "lda") {
    mod <- lda(y ~ ., data = df_tr)
  } else {
    mod <- qda(y ~ ., data = df_tr)
  }

  pred_tr <- predict(mod, df_tr)$class
  pred_te <- predict(mod, df_te)$class

  c(train_error = mean(pred_tr != ytr),
    test_error  = mean(pred_te != yte))
}
```

```{r da-fit}
lda_std_res <- disc_analysis("lda", X_train_std, y_train_fac, X_test_std, y_test_fac)
qda_std_res <- disc_analysis("qda", X_train_std, y_train_fac, X_test_std, y_test_fac)
lda_log_res <- disc_analysis("lda", X_train_log, y_train_fac, X_test_log, y_test_fac)
qda_log_res <- disc_analysis("qda", X_train_log, y_train_fac, X_test_log, y_test_fac)

da_errs <- data.frame(
  Method = c("LDA (Standardized)", "QDA (Standardized)",
             "LDA (Log-transformed)", "QDA (Log-transformed)"),
  Train = c(lda_std_res["train_error"], qda_std_res["train_error"],
            lda_log_res["train_error"], qda_log_res["train_error"]),
  Test  = c(lda_std_res["test_error"], qda_std_res["test_error"],
            lda_log_res["test_error"], qda_log_res["test_error"])
)
knitr::kable(da_errs, digits = 3, col.names = c("Method", "Training Error", "Test Error"))
```

QDA does noticeably worse than LDA here -- probably overfitting because
it has to estimate a full 57×57 covariance matrix per class and there
aren't enough samples to do that reliably.

---

## (d) SVM

Testing linear and RBF kernels on all three preprocessed versions.

```{r svm-helper}
fit_svm <- function(kern, xtr, ytr, xte, yte) {
  df_tr <- data.frame(xtr, y = ytr)
  df_te <- data.frame(xte, y = yte)

  mod <- svm(y ~ ., data = df_tr,
             kernel = kern,
             cost = 1,
             gamma = 1 / ncol(xtr),
             scale = FALSE)

  p_tr <- predict(mod, df_tr)
  p_te <- predict(mod, df_te)

  c(train_error = mean(p_tr != ytr),
    test_error  = mean(p_te != yte))
}
```

```{r svm-fit}
data_versions <- list(
  Std = list(tr = X_train_std, te = X_test_std),
  Log = list(tr = X_train_log, te = X_test_log),
  Bin = list(tr = X_train_bin, te = X_test_bin)
)

svm_res <- do.call(rbind, lapply(names(data_versions), function(v) {
  d <- data_versions[[v]]
  lin <- fit_svm("linear", d$tr, y_train_fac, d$te, y_test_fac)
  rbf <- fit_svm("radial", d$tr, y_train_fac, d$te, y_test_fac)

  data.frame(
    Preprocessing = v,
    Kernel = c("Linear", "RBF"),
    Train  = c(lin["train_error"], rbf["train_error"]),
    Test   = c(lin["test_error"],  rbf["test_error"])
  )
}))

knitr::kable(svm_res, digits = 3,
             col.names = c("Preprocessing", "Kernel", "Train Err", "Test Err"))
```

---

## All results

```{r combined-table}
# clean up svm results for merging
svm_fmt <- svm_res
svm_fmt$Method <- paste("SVM", svm_fmt$Kernel)
svm_fmt$Data   <- svm_fmt$Preprocessing
svm_fmt <- svm_fmt[, c("Method", "Data", "Train", "Test")]

everything <- rbind(
  data.frame(Method = "Logistic", Data = "Standardized",
             Train = lr_std$train_err, Test = lr_std$test_err),
  data.frame(Method = "Logistic", Data = "Log",
             Train = lr_log$train_err, Test = lr_log$test_err),
  data.frame(Method = "Logistic", Data = "Binarized",
             Train = lr_bin$train_err, Test = lr_bin$test_err),
  data.frame(Method = "LDA", Data = "Standardized",
             Train = as.numeric(lda_std_res["train_error"]),
             Test  = as.numeric(lda_std_res["test_error"])),
  data.frame(Method = "QDA", Data = "Standardized",
             Train = as.numeric(qda_std_res["train_error"]),
             Test  = as.numeric(qda_std_res["test_error"])),
  data.frame(Method = "LDA", Data = "Log",
             Train = as.numeric(lda_log_res["train_error"]),
             Test  = as.numeric(lda_log_res["test_error"])),
  data.frame(Method = "QDA", Data = "Log",
             Train = as.numeric(qda_log_res["train_error"]),
             Test  = as.numeric(qda_log_res["test_error"])),
  svm_fmt
)

knitr::kable(everything, digits = 3,
             col.names = c("Method", "Data", "Train Error", "Test Error"))
```

### Conclusions

The linear SVM on log-transformed data gave the best test error at 0.056,
with the RBF SVM on the same data right behind at ~0.057. The fact that
a linear kernel does this well tells me the classes are mostly linearly
separable once the features are log-transformed. Train error (0.058) and
test error (0.056) are close, so the model isn't overfitting -- that's
reassuring.

Still, ~5% misclassification means roughly 1 in 20 emails gets the wrong
label. There's room for improvement with hyperparameter tuning, ensembles,
or better feature engineering.
