---
title: "Spam Email Classification"
author: "Aneesh Dalvi"
date: "2025-06-07"
output:
  html_document:
    toc: true
    toc_float: true
---

## Loading the data

```{r}
# read in the spam dataset -- no headers, comma-separated
train <- read.table("spam-train.txt", sep = ",", header = FALSE)
test  <- read.table("spam-test.txt",  sep = ",", header = FALSE)

# first 57 cols are features, col 58 is the label
train_x <- train[, 1:57]
train_y <- train[, 58]
test_x  <- test[, 1:57]
test_y  <- test[, 58]

# gonna need these as factors later
train_labels <- factor(train_y, levels = c(0, 1), labels = c("ham", "spam"))
test_labels  <- factor(test_y,  levels = c(0, 1), labels = c("ham", "spam"))
```

## Preprocessing

Three different transformations on the raw features.

### 1) Standardization (zero mean, unit variance)

```{r}
# compute mean/sd from training set only to avoid data leakage
mu  <- apply(train_x, 2, mean)
sig <- apply(train_x, 2, sd)

train_std <- scale(train_x, center = mu, scale = sig)
test_std  <- scale(test_x,  center = mu, scale = sig)
```

### 2) Log transform

```{r}
# log(x + 1) to handle zeros
train_log <- log(train_x + 1)
test_log  <- log(test_x + 1)
```

### 3) Binarize (presence/absence)

```{r}
# just indicator of whether the feature is nonzero
train_bin <- as.data.frame((train_x > 0) * 1)
test_bin  <- as.data.frame((test_x > 0) * 1)
```

---

## Part (a): PCA Visualization

Using PCA to get a 2D view of each preprocessed version of the data.

```{r fig.width=14, fig.height=5}
library(ggplot2)
library(gridExtra)

# --- standardized ---
pc_std <- prcomp(train_std, center = FALSE, scale. = FALSE)
plot_std <- data.frame(PC1 = pc_std$x[, 1], PC2 = pc_std$x[, 2], Class = train_labels)

g1 <- ggplot(plot_std, aes(PC1, PC2, color = Class)) +
  geom_point(alpha = 0.6, size = 1) +
  ggtitle("Standardized") +
  theme_minimal()

# --- log-transformed ---
pc_log <- prcomp(train_log, center = TRUE, scale. = TRUE)
plot_log <- data.frame(PC1 = pc_log$x[, 1], PC2 = pc_log$x[, 2], Class = train_labels)

g2 <- ggplot(plot_log, aes(PC1, PC2, color = Class)) +
  geom_point(alpha = 0.6, size = 1) +
  ggtitle("Log-transformed") +
  theme_minimal()

# --- binarized ---
pc_bin <- prcomp(as.data.frame(lapply(train_bin, as.numeric)), center = TRUE, scale. = FALSE)
plot_bin <- data.frame(PC1 = pc_bin$x[, 1], PC2 = pc_bin$x[, 2], Class = train_labels)

g3 <- ggplot(plot_bin, aes(PC1, PC2, color = Class)) +
  geom_point(alpha = 0.6, size = 1) +
  ggtitle("Binarized") +
  theme_minimal()

grid.arrange(g1, g2, g3, ncol = 3)
```

---

## Part (b): Logistic Regression

Fitting logistic regression on each version. I also check which features
are significant after Bonferroni correction since there are 57 predictors
(multiple testing).

```{r}
# bonferroni threshold -- 57 tests
alpha_bonf <- 0.05 / ncol(train_std)

fit_logistic <- function(xtr, ytr, xte, yte) {
  # stick everything into data frames for glm
  df_train <- data.frame(xtr, spam = ytr)
  df_test  <- data.frame(xte, spam = yte)

  mod <- glm(spam ~ ., data = df_train, family = binomial)

  # predictions
  pred_train <- predict(mod, type = "response") > 0.5
  pred_test  <- predict(mod, newdata = df_test, type = "response") > 0.5

  # misclassification rates
  err_train <- mean(pred_train != (ytr == "spam"))
  err_test  <- mean(pred_test  != (yte == "spam"))

  # grab p-values from the coefficient table
  coef_table <- summary(mod)$coefficients
  pv <- coef_table[, 4]
  sig_raw  <- names(pv)[pv < 0.05]
  sig_bonf <- names(pv)[pv < alpha_bonf]

  list(
    tr_err   = err_train,
    te_err   = err_test,
    sig_raw  = sig_raw,
    sig_bonf = sig_bonf,
    coefs    = coef_table
  )
}

lr_std <- fit_logistic(train_std, train_labels, test_std, test_labels)
lr_log <- fit_logistic(train_log, train_labels, test_log, test_labels)
lr_bin <- fit_logistic(train_bin, train_labels, test_bin, test_labels)
```

### Error rates

```{r}
lr_table <- data.frame(
  Preprocessing = c("Standardized", "Log-transformed", "Binarized"),
  Train_Error   = c(lr_std$tr_err, lr_log$tr_err, lr_bin$tr_err),
  Test_Error    = c(lr_std$te_err, lr_log$te_err, lr_bin$te_err)
)
knitr::kable(lr_table, digits = 3, col.names = c("Preprocessing", "Train Err", "Test Err"))
```

### Significant features

```{r}
cat("Standardized -- raw p < 0.05:\n",
    paste(lr_std$sig_raw, collapse = ", "), "\n\n")

cat("Standardized -- Bonferroni corrected:\n",
    paste(lr_std$sig_bonf, collapse = ", "), "\n\n")

cat("Log-transformed -- raw p < 0.05:\n",
    paste(lr_log$sig_raw, collapse = ", "), "\n\n")

cat("Log-transformed -- Bonferroni corrected:\n",
    paste(lr_log$sig_bonf, collapse = ", "), "\n\n")

cat("Binarized -- raw p < 0.05:\n",
    paste(lr_bin$sig_raw, collapse = ", "), "\n\n")

cat("Binarized -- Bonferroni corrected:\n",
    paste(lr_bin$sig_bonf, collapse = ", "), "\n")
```

The log-transformed version gave the best test error (~5.7%), then standardized (~7.3%),
and binarized came in last (~8.1%). After applying Bonferroni correction (threshold ≈ 0.00088),
about a dozen features stay significant across all three models -- particularly V5, V7, V16,
V17, V21, V23, V25, V27, V45, V46, V52, and V53. These are likely the strongest spam
indicators (certain word frequencies and capital letter statistics). The binarized model
picks up a few extras (V11, V13, V24, V28, V37, V42), which makes sense because for a lot
of tokens, just knowing whether they appear at all is almost as useful as knowing the
exact count.

---

## Part (c): LDA and QDA

Running discriminant analysis on the standardized and log-transformed data.
(Skipping binarized here since binary features don't play nice with QDA's
covariance estimation.)

```{r}
library(MASS)

run_da <- function(type, xtr, ytr, xte, yte) {
  df_tr <- data.frame(xtr, spam = ytr)
  df_te <- data.frame(xte, spam = yte)

  if (type == "lda") {
    model <- lda(spam ~ ., data = df_tr)
  } else {
    model <- qda(spam ~ ., data = df_tr)
  }

  p_tr <- predict(model, df_tr)$class
  p_te <- predict(model, df_te)$class

  c(train_error = mean(p_tr != ytr),
    test_error  = mean(p_te != yte))
}

lda_std <- run_da("lda", train_std, train_labels, test_std, test_labels)
qda_std <- run_da("qda", train_std, train_labels, test_std, test_labels)
lda_log <- run_da("lda", train_log, train_labels, test_log, test_labels)
qda_log <- run_da("qda", train_log, train_labels, test_log, test_labels)

da_table <- data.frame(
  Method      = c("LDA (Std)", "QDA (Std)", "LDA (Log)", "QDA (Log)"),
  Train_Error = c(lda_std["train_error"], qda_std["train_error"],
                  lda_log["train_error"], qda_log["train_error"]),
  Test_Error  = c(lda_std["test_error"], qda_std["test_error"],
                  lda_log["test_error"], qda_log["test_error"])
)
knitr::kable(da_table, digits = 3, col.names = c("Method", "Training Error", "Test Error"))
```

---

## Part (d): SVM (Linear + RBF)

Trying both linear and radial basis function kernels on all three
preprocessed versions.

```{r}
library(e1071)

run_svm <- function(kern, xtr, ytr, xte, yte) {
  df_tr <- data.frame(xtr, spam = ytr)
  df_te <- data.frame(xte, spam = yte)

  fit <- svm(spam ~ ., data = df_tr,
             kernel = kern,
             cost = 1,
             gamma = 1 / ncol(xtr),
             scale = FALSE)

  ptr <- predict(fit, df_tr)
  pte <- predict(fit, df_te)

  c(train_error = mean(ptr != ytr),
    test_error  = mean(pte != yte))
}

# run all combos
datasets <- list(
  Std = list(tr = train_std, te = test_std),
  Log = list(tr = train_log, te = test_log),
  Bin = list(tr = train_bin, te = test_bin)
)

svm_results <- do.call(rbind, lapply(names(datasets), function(nm) {
  d <- datasets[[nm]]
  lin <- run_svm("linear", d$tr, train_labels, d$te, test_labels)
  rbf <- run_svm("radial", d$tr, train_labels, d$te, test_labels)
  data.frame(
    Preprocessing = nm,
    Kernel        = c("Linear", "RBF"),
    Train_Error   = c(lin["train_error"], rbf["train_error"]),
    Test_Error    = c(lin["test_error"],  rbf["test_error"])
  )
}))

knitr::kable(svm_results, digits = 3,
             col.names = c("Preprocessing", "Kernel", "Train Err", "Test Err"))
```

---

## Summary Table

Putting everything together in one place.

```{r}
# build a clean combined table
svm_fmt <- svm_results
svm_fmt$Method <- paste("SVM", svm_fmt$Kernel)
svm_fmt$Data   <- svm_fmt$Preprocessing
svm_fmt$Train  <- svm_fmt$Train_Error
svm_fmt$Test   <- svm_fmt$Test_Error
svm_fmt <- svm_fmt[, c("Method", "Data", "Train", "Test")]

all_results <- rbind(
  data.frame(Method = "Logistic", Data = "Standardized",
             Train = lr_std$tr_err, Test = lr_std$te_err),
  data.frame(Method = "Logistic", Data = "Log",
             Train = lr_log$tr_err, Test = lr_log$te_err),
  data.frame(Method = "Logistic", Data = "Binarized",
             Train = lr_bin$tr_err, Test = lr_bin$te_err),
  data.frame(Method = "LDA", Data = "Standardized",
             Train = as.numeric(lda_std["train_error"]),
             Test  = as.numeric(lda_std["test_error"])),
  data.frame(Method = "QDA", Data = "Standardized",
             Train = as.numeric(qda_std["train_error"]),
             Test  = as.numeric(qda_std["test_error"])),
  data.frame(Method = "LDA", Data = "Log",
             Train = as.numeric(lda_log["train_error"]),
             Test  = as.numeric(lda_log["test_error"])),
  data.frame(Method = "QDA", Data = "Log",
             Train = as.numeric(qda_log["train_error"]),
             Test  = as.numeric(qda_log["test_error"])),
  svm_fmt
)

knitr::kable(all_results, digits = 3,
             col.names = c("Method", "Data", "Train Error", "Test Error"))
```

### Takeaways

The linear SVM on log-transformed data gave the lowest test error at about 5.6%,
with the RBF SVM on the same data right behind it at ~5.7%. The fact that a linear
kernel does this well tells me the classes are mostly linearly separable in the
log-transformed feature space. The train and test errors are close to each other
(~5.8% train vs ~5.6% test), so the model isn't overfitting, which is good.

That said, ~5% error still means roughly 1 in 20 emails gets misclassified.
There's definitely room for improvement -- hyperparameter tuning, ensemble
methods, or more sophisticated feature engineering could probably push this lower.
